---
title: "Cross-feeding of two gastrointestinal bacteria"
author: "Silvio Waschina"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crossfeeding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(Eutropia)
```

## Background
The intestinal bacterium *Eubacterium rectale* is known to be able to use acetate as energy source under anaerobic conditions and thereby forms butyrate as end product (Rivère et al. (2015) Appl Envrion Microbiol). Acetate is a common fermentation end product in a number of different other intestinal bacteria, including Bifidobacteria (e.g. *Bifidobacterium longum*). In this tutorial, a co-culture experiment of both species is simulated in an aerobic environment that contains glucose as sole source of carbon an energy. It will be tested if acetate cross-feeding occurs and how the organism's affect each others population growth.

## The setup

As growth environment, a Petri dish-shaped growth area will be defined with a radius of 80 micro meters. The space for metabolite concentrations is rasterized by rhombic dodecahedrons of size 1 µm (distance between adjacent field centers). The chemical growth environment consists of 8 field layers in height.
```{r}
# create simulation environment
sim <- init_simulation("Petri_80",
                       gridFieldSize = 1,
                       gridFieldLayers = 8)
```

Next, we will read the models for both organisms and place 15 cells each in the growth environment.
```{r}
# add organisms (15 cells of each as 'starter culture')
models <- list()
models[['eure']] <- readRDS(system.file("extdata", "eure.RDS", package="Eutropia"))
models[['bilo']] <- readRDS(system.file("extdata", "bilo.RDS", package="Eutropia"))

sim <- add_organism(sim, model = models[['eure']], name = "E. rectale",
                    ncells = 15, distribution.radius = 25,
                    open.bounds = 1)
sim <- add_organism(sim, model = models[['bilo']], name = "B. longum",
                    ncells = 15, distribution.radius = 25,
                    open.bounds = 1)
```

It is not required, but out of curiosity the initial spatial distribution of cells can be plotted.

```{r fig.height=5}
plot_cells(sim,
           scalebar.color = "black")
```

Now we have an simulation object with cells. What so far is missing are nutrients, or a growth medium.

```{r}
# Load a medium table
dt_medium <- fread(system.file("extdata", "medium.csv", package="Eutropia"))

# adding compounds
sim <- add_compounds(sim,
                     compounds = dt_medium$cpd.id,
                     concentrations = dt_medium$mM,
                     compound.names = dt_medium$cpd.name,
                     is.constant = dt_medium$is.constant)
```

## The simulation

We are all set to run our first simulation. It will run for 50 iterations (= 500 minutes simulated time). The simulation may take around 4 minutes 30 seconds to finish, because of the option `lim_time = 4.5`, while the last iteration is allowed to finish even if the time limit is exceeded. You can of course increase the time limit. The number of iterations depends on the systems performance.

```{r}
sim <- run_simulation(sim, niter = 50, verbose = 1, lim_time = 4.5)
```

## Analyse the environment after the simulation

The package comes with a number of functions to plot simulation results.

```{r fig.height=5}
# Plot spatial distribution of cells
plot_cells(sim, scalebar.color = "black")
```

```{r}
# Spatial distribution of glucose
plot_environment(sim, compounds = c("cpd00027_e0"),
                 scalebar.color = "black")
```

```{r fig.height=3, fig.width=9}
# Spatial distribution of acetate, lactate and butyrate
plot_environment(sim,
                 compounds = c("cpd00029_e0","cpd00211_e0","cpd00159_e0"),
                 scalebar.color = "black")
```

```{r fig.width=5}
# growth curves (total cell mass by species)
plot_growth(sim)
```

```{r fig.width=5}
# dynamics in metabolite concentrations (calculated by total environment volume)
plot_compounds(sim, compounds = c("cpd00029_e0","cpd00211_e0","cpd00159_e0"))
```

The consumed and produced metabolites can be retrieved for the simulations.
```{r}
summary_exchanges(sim)
```

In addition, the same information can be retrieved for earlier iterations in the simulation.
```{r}
summary_exchanges(sim, iter = 10)
```

The results should clearly show that, while acetate is produced by *B. longum*, it is consumed by *E. rectale*.
