---
title: "Cross-feeding of two gastrointestinal bacteria"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Crossfeeding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(EcoAgents)
```

## Background
The intestinal bacterium *Eubacterium rectale* is known to be able to use acetate as energy source under anaerobic conditions and thereby forms butyrate as end product (Rivère et al. (2015) Appl Envrion Microbiol). Acetate is a common fermentation end product in a number of different other intestinal bacteria, including Bifidobacteria (e.g. *Bifidobacterium longum*). In this tutorial, a co-culture experiment of both species is simulated in an aerobic environment that contains glucose as sole source of carbon an energy. It will be tested if acetate cross-feeding occurs and how the organism's affect each others population growth.

## The setup

As growth environment, a Petri dish-shaped growth area will be defined with a radius of 80 micro meters. The space metabolite concentrations is rasterized by rhombic dodecahedrons of size 1 µm (distance between field centers). The chemical growth environment consists of 8 field layers in height.
```{r}
# create simulation environment
sim <- init_simulation("Petri_80",
                       gridFieldSize = 1,
                       gridFieldLayers = 8)
```

Next, we will read the models for both organisms and place 15 cells each in the growth environment.
```{r}

# add organisms (15 cells of each as 'starter culture')
models <- list()
models[['eure']] <- readRDS(system.file("extdata", "eure.RDS", package="EcoAgents"))
models[['bilo']] <- readRDS(system.file("extdata", "bilo.RDS", package="EcoAgents"))

sim <- add_organism(sim, model = models[['eure']], name = "E. rectale",
                    ncells = 15, distribution.radius = 25,
                    open.bounds = 1)
sim <- add_organism(sim, model = models[['bilo']], name = "B. longum",
                    ncells = 15, distribution.radius = 25,
                    open.bounds = 1)
```

It is not required, but out of curiosity the intial spatial distribution of cells can be plotted

```{r fig.height=5}
plot_cells(sim,
           scalebar.color = "black")
```

Now we have an simulation object with cells. What so far missing are nutrients, or a growth medium.

```{r}
# Load a medium table
dt_medium <- fread(system.file("extdata", "medium.csv", package="EcoAgents"))

# adding compounds
sim <- add_compounds(sim,
                     compounds = dt_medium$cpd.id,
                     concentrations = dt_medium$mM,
                     compound.names = dt_medium$cpd.name,
                     is.constant = dt_medium$is.constant)
```

## The simulation

We are all set to run our first simulation. It will run for 50 iterations (= 500 minutes simulated time). The simulation may take between 5 to 10 minutes to finish.

```{r}
sim <- run_simulation(sim, niter = 50, verbose = 1)
```

## Analyse the environment after the simulation

The package comes with a number of function to plot simulation results.

```{r fig.height=5}
# Plot spatial distribution of cells
plot_cells(sim,
           scalebar.color = "black")
```

```{r}
# Spatial distribution of glucose
plot_environment(sim, compounds = c("cpd00027_e0"),
                 scalebar.color = "black")
```

```{r fig.height=3, fig.width=9}
# Spatial distribution of acetate, lactate and butyrate
plot_environment(sim,
                 compounds = c("cpd00029_e0","cpd00211_e0","cpd00159_e0"),
                 scalebar.color = "black")
```

```{r fig.width=5}
# growth curves (total cell mass by species)
plot_growth(sim)
```

```{r fig.width=5}
# dynamics in metabolite concentrations (calculated by total environment volume)
plot_compounds(sim, compounds = c("cpd00029_e0","cpd00211_e0","cpd00159_e0"))
```

The consumed and produced metabolites can be retrieved for the simulations.

```{r}
summary_exchanges(sim)
```

In addition, the same information can be retrieved for earlier iterations in the simulation.
```{r}
summary_exchanges(sim, iter = 25)
```

The results should clearly show that, while acetate is produced by *B. longum*, it is consumed by *E. rectale*.
